Switching Optimisation in Huffman Code for Power Efficient Data Transmission

### 2015-02-27 Switche Paper 1st approache/elsarticle-template-num.tex
some work

The rest of the paper is organised as follows:
Section 2 presents the background study of the
relationship of switching activity with power consumption, the relation of switching with Huffman code and the issues in biological data transmission. The proposed approach is described in Section 3. Experimental results and
discussion are presented in Section 4 . Finally,
concluding remarks are presented in Section 5. 
\section{Background}
\subsection{Effect of Switching in Power Consumption}
CMOS is the most widely used technology implemented in VLSI chips because of their power efficiency. Power dissipation in CMOS circuits has three different components: dynamic, static, and leakage \cite{Weste88}.
\begin{equation*}
P_{avg} = P_{dynamic}+P_{static}+P_{leakge} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\end{equation*}
\begin{equation}
~~~~~~~~~~~~= 0.5 \times C_{L} \times V_{dd}^{2}\times E(sw)\times f_{clk} + I_{sc} \times V_{dd}+I_{leakage} \times V_{dd}
 \end{equation}
 where\\
  $C_{L}$ is the load capacitance, \\
  $V_{dd}$ is power supply,\\
  $E(sw)$ is average transition (switching) number per clock cycle ,\\
  $f_{clk}$ is the clock frequency,\\
  $ I_{sc}$ is the short circuit current, and\\
  $ I_{leakage}$ is the leakage current.\\
Out of the three components, the dynamic power dissipation is the dominant and it counts for 70\%-90\% of the total power consumption\cite{pedram1996}. Therefore, optimisation of dynamic power consumption is considered in this paper. In CMOS circuit, dynamic power consumption arises when the load capacitance ($C_{L}$) is charged to address a transition from low to the high voltage level.  The parameters $C_{L}$, $V_{dd}$, and $f_{clk}$ in dynamic power dissipation are primarily determined by the circuit layout and the fabrication technology. However, amount of switching, $E(sw)$ completely depends on the digital representation of the data. As the dynamic power dissipation is proportional  to the average number of switching, thus a reduction in the total switching number will eventually results in the reduction of the dynamic power consumption. Therefore, in this paper, we focus on reducing total amount of switching in Huffman code to improve power efficiency of digital data processing.
%In this paper, by switching we mean a transition from `0' to `1' or `1' to `0' in a binary sequence. 
\subsection{Huffman Code and Its Relationship with Switching Activity}
%\subsubsection{Relationship between switching and Huffman Code}
In computer science and information theory, Huffman code is an entropy encoding algorithm used for lossless data compression. The classical Huffman code requires to pass the input data two times to complete the encoding operation. In the first pass, it reads the whole stream to determine the occurrence frequencies of distinct symbols. After that, symbols along with their frequencies are used to generate a binary tree known as Huffman tree. The process of Huffman tree generation is shown in algorithm 1 \cite{Cormen2001}.

\begin{algorithm}[!thpb]
\label{alg1}
\caption{Huffman(C)}
\begin{algorithmic}[1]
\State $n \leftarrow |C|$
\State $Q \leftarrow C$
\FOR {$i = 1~to~n-1$}
\State $allocate~a~new~node~z$
\State $left[z]$ $\leftarrow$ $x$ $\leftarrow$ $EXTRACT\_MIN (Q)$
\State $right[z]$ $\leftarrow$ $y$ $\leftarrow$ $EXTRACT\_MIN (Q)$
\State $f[z]$ $\leftarrow$ $f[x] + f[y]$
\State $INSERT (Q~,~z)$
\ENDFOR
\State \textbf{$retur$n} $EXTRACT-MIN (Q)$
\end{algorithmic}
\end{algorithm}
In the algorithm, C is the set of n distinct symbols, and each symbol $ c\in C$ has a frequency denoted as f[c]. The algorithm starts its operation with a set of $|C|$  leaf nodes where each node corresponds to a distinct symbol. After that, a sequence of $|C|-1$ merging operation is performed to obtain the final tree. The merging is done by extracting two least frequent symbols from the minimum priority queue, $Q$, and creating and inserting a new node (parent) to the queue for that two symbols. The frequency of the new  node is calculated as the sum of the frequency of its child nodes.  The resultant tree ensures the optimal encoding of symbols to obtain maximal compression performance. The tree is traversed from the root to the leaf nodes to obtain the unique codeword (binary string) for different symbols. Once the codewords are obtained, in the second pass, the Huffman algorithm encodes the symbols of the input message with the codewords, thus reduce the size of the whole message.
For a set of symbols more than one Huffman tree is possible, i.e., optimal solution can be obtained from different structure of the tree. If the structure of the tree is changed, the codewords for the symbols will also change. The switching activity depends on the structure of the codewords. For example, if a symbol is assigned a codewrod `000' then there is no switching, on the other hand, if the codeword `010' is assigned to a symbol the transmission of this symbol requires 2 switching activity.

\subsection{Issues in Biological data transmission}
The size of biological data base increase with an expanding rate and it doesn't stop for increasing. In the last decade the use of biological data history increase which evolve a high request to data from this high volume biological data base~\cite{Doug08}. The exponential growth of these databases become a big problem to all biological data processing methods~\cite{Doug08}. Data compression methods is the most used methods to evolve the reduce the cost of treatment of high volume data bases. The main objective of data compression methods is minimising the number of bits in the data representation. Authors in~\cite{bra09} proposed a new approach to genomic data compression based in suffix and common substring and code the difference with Huffman code. In~\cite{chr09}, authors propose a general scheme for coding only the difference between the target genome and the referential one, the coding method used is Huffman. Recent work on data compression for biological data have been proposed to deal with transmission of genomic data as an email attachment~\cite{wan11}. Cost of data transmission is based on two point, firstly the size of the data which has been improved considerably in the recent year with the improvement of Huffman code. Secondly, is the number of switches on the generated codes. The power consumption increase linear with the number of switches.
\section{GA-SO : Genetic algorithm for Switches Optimising}
Genetic algorithm (GA) is one the of the most popular bio-inspired meta-heuristic algorithm inspired from the natural evolution of species~\cite{gen1}. It is a  population based algorithm starts with the generation of a random initial population. The population contain a set of feasible solutions called individuals, that are usually far from the optimal. The GA optimisation process uses a set of natural genetic operators such as selection, crossover and mutation to converge to the optimal solution.
\subsection{Preparing data and population generation}
A genome is an expansion of DNA sequence which is a set of nucleotides that encodes a protein.These nucleotide are bounded together and each order define a specific protein. The transformation of DNA sequence to protein  is made by the cellular machinery, each three nucleotides are interpreted to a specific amino acid this three nucleotide are called triplets~\cite{Harvey00}. Huffman algorithm read the whole message to compute frequencies, the proposed algorithm read the whole genome and cut it into triplet to generate the frequencies of the triplet and by the way to generate the adjacent matrix which represents the frequencies precedence between triplet. The initial population is generated randomly, each solution on the population represents a tree that generates a set of codes equal to the total triplets number.
\subsection{Selection}

@@

\subsection{Crossover}

@@

\subsection{Mutation}

>>>>

The rest of the paper is organised as follows:
Section 2 presents the background study of the
relationship of switching activity with power consumption, the relation of switching with Huffman code and the issues in biological data transmission. The proposed approach is described in Section 3. Experimental results and
discussion are presented in Section 4 . Finally,
concluding remarks are presented in Section 5. 
\section{Background}
\subsection{Effect of Switching in Power Consumption}
CMOS is the most widely used technology implemented in VLSI chips because of their power efficiency. Power dissipation in CMOS circuits has three different components: dynamic, static, and leakage \cite{Weste88}.
\begin{equation*}
P_{avg} = P_{dynamic}+P_{static}+P_{leakge} ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
\end{equation*}
\begin{equation}
~~~~~~~~~~~~= 0.5 \times C_{L} \times V_{dd}^{2}\times E(sw)\times f_{clk} + I_{sc} \times V_{dd}+I_{leakage} \times V_{dd}
 \end{equation}
 where\\
  $P_{avg}$ is the average power dissipation
  $C_{L}$ is the load capacitance, \\
  $V_{dd}$ is power supply,\\
  $E(sw)$ is average transition (switching) number per clock cycle ,\\
  $f_{clk}$ is the clock frequency,\\
  $ I_{sc}$ is the short circuit current, and\\
  $ I_{leakage}$ is the leakage current.\\
Out of the three components, the dynamic power dissipation is the dominant and it counts for 70\%-90\% of the total power consumption\cite{pedram1996}. Therefore, optimisation of dynamic power consumption is considered in this paper. In CMOS circuit, dynamic power consumption arises when the load capacitance ($C_{L}$) is charged to address a transition from low to the high voltage level.  The parameters $C_{L}$, $V_{dd}$, and $f_{clk}$ in dynamic power dissipation are primarily determined by the circuit layout and the fabrication technology. However, amount of switching, $E(sw)$ completely depends on the digital representation of the data. As the dynamic power dissipation is proportional  to the average number of switching, thus a reduction in the total switching number will eventually results in the reduction of the dynamic power consumption. Therefore, we focus on reducing total amount of switching in Huffman code to improve power efficiency of digital data processing.
%In this paper, by switching we mean a transition from `0' to `1' or `1' to `0' in a binary sequence. 
\subsection{Huffman Code and Its Relationship with Switching Activity}
%\subsubsection{Relationship between switching and Huffman Code}
In computer science and information theory, Huffman code is an entropy encoding algorithm used for lossless data compression. The classical Huffman code requires to pass the input data two times to complete the encoding operation. In the first pass, it reads the whole stream to determine the occurrence frequencies of distinct symbols. After that, symbols along with their frequencies are used to generate a binary tree known as Huffman tree. The process of Huffman tree generation is shown in algorithm 1 \cite{Cormen2001}.

\begin{algorithm}[!thpb]
\label{alg1}
\caption{Huffman(C)}
\begin{algorithmic}[1]
\State $n \leftarrow |C|$
\State $Q \leftarrow C$
\FOR {$i = 1~to~n-1$}
\State $allocate~a~new~node~z$
\State $left[z]$ $\leftarrow$ $x$ $\leftarrow$ $EXTRACT\_MIN (Q)$
\State $right[z]$ $\leftarrow$ $y$ $\leftarrow$ $EXTRACT\_MIN (Q)$
\State $f[z]$ $\leftarrow$ $f[x] + f[y]$
\State $INSERT (Q~,~z)$
\ENDFOR
\State \textbf{$retur$n} $EXTRACT\_MIN (Q)$
\end{algorithmic}
\end{algorithm}
In the algorithm, C is the set of n distinct symbols, and each symbol $ c\in C$ has a frequency denoted as f[c]. The algorithm starts its operation with a set of $|C|$  leaf nodes where each node corresponds to a distinct symbol. After that, a sequence of $|C|-1$ merging operation is performed to obtain the final tree. The merging is done by extracting two least frequent symbols from the minimum priority queue, $Q$, and creating and inserting a new node (parent) to the queue for that two symbols. These two symbols are considered as the left and right child of their parent node respectively. The frequency of the new  node is calculated as the sum of the frequency of its child nodes.  The resultant tree ensures the optimal encoding of symbols to obtain maximal compression performance. %The tree is traversed from the root to the leaf nodes to obtain the unique codeword (binary string) for different symbols. 
To form a codeword for a symbol, the tree is traversed from the  root to the leaf node representing the symbol. In the traversal process, any move made towards the left direction is replaced by `0' and any move towards the right direction is replaced by `1'. As a result codeword for each symbol consist of a sequence of 0's and/or 1's. Once the codewords are obtained, in the second pass, the Huffman algorithm encodes the symbols of the input message with the codewords, thus reduce the size of the whole message.
For a set of symbols more than one Huffman tree is possible, i.e., optimal solution can be obtained from different structure of the tree. As the codeword of a symbol depend on its position in the tree, therefore changing the structure of the tree will change the codeword for the symbol. A codeword can contain zero or more switching activity within itself refereed to as intra-switch. For example, if a symbol `A' is assigned a codewrod `000' then there is no intra-switch, on the other hand, if the codeword `101' is assigned to a symbol `B` then their exist two intra-switches. When the codewords of different symbols are concatenated together during transmission they form switching based on the head and tail bit of the codewords. For example, if we want to transmit `AB' then according to the above codewords we will transmit `00\textbf{01}01' which will result in an inter-switch because the tail bit (0) of `A' and the head bit (1) of `B' are different. That means, in this case, we have three total switches--two intra-switches and one inter-switch. Now if `B' is assigned the codeword `010', the new transmitted string is `00\textbf{00}10', then we still have two intra-switches for the codeword of `B' but the inter-switch no more exists. So, changing tree structure, i.e., codewords, will change the total number of switching. Hence, the aim of this paper is to find an optimal Huffman tree that will minimise the total number of switching by making a balance between intra- and inter-switches.

\subsection{Issues in Biological data transmission}
The size of biological databases are continuously increasing. In the last decade the access to these databases has increased significantly. The exponential growth of these databases pose a big problem to all biological data processing methods~\cite{Doug08}. Data compression methods are widely used to reduce the overheads of treating, e.g., transmitting, the high volume of data. The main objective of data compression methods is minimising the number of bits in the data representation. Authors in~\cite{bra09} proposed a new approach to genomic data compression based in suffix and common substring and code the difference with Huffman code. In~\cite{chr09}, authors propose a general scheme based on Huffman code for coding only the difference between the target genome and the referential one. Recently, a method has been proposed to deal with the transmission of genomic data as an email attachment~\cite{wan11}. All the methods used for biological data compression aims only at reducing the size of the data. However, due to the huge size of the input data, the compressed data contains a huge number of switching in them which has a big impact on power consumption. As no consideration is paid to the reduction of switching, hence, the compression process may be size efficient but not power efficient. Therefore, it is believed that a more efficient compression of biological data can be obtained by optimising the size of the data as well as the total number of switching in the transmitted data. %Cost of data transmission is based on two point, firstly the size of the data which has been improved considerably in the recent year with the improvement of Huffman code. Secondly, is the number of switches on the generated codes. %The power consumption increase linear with the number of switches.
\section{The proposed approach for Switching Optimisation}
The main objective of the approach is to minimise the total number of switching in the representation of data produced by the Huffman algorithm. To do so, it is required to minimise both intra- and inter-switches among the codewords. The total number of intra-switch depends on the codewords themselves and the frequency of the symbols. As a result, for  a given set of codewords an optimal number of intra-switches can be obtained by assigning the most frequent symbol with the codewords with minimum number of switching. However, it will not ensure the optimal number of inter-switches. Because, inter-switches not only depend on the codewords of the symbols but also the association between the symbols. By association, we mean how many times a symbol occurs after another symbol in the whole message, i.e., all possible pairing combination among the symbols. As the number of distinct symbols increases in the message the number of combination increases. So improving total number of inter-switching is a combinatorial problem.

Genetic algorithm (GA) has widely been applied to solve a range of combinatorial problem in complex VLSI design~\cite{chan1991macro,bright1996genetic,cohoon1987genetic}. For this reason, the GA is identified as a potential solution to the above mentioned problem. GA is one the of the most popular bio-inspired meta-heuristic algorithm inspired by the natural evolution of species~\cite{gen1}. It is a population based algorithm starts with the generation of a random initial population. The population contain a set of feasible solutions called individuals, that are usually far from the optimal. The GA optimisation process uses a set of natural genetic operators such as selection, crossover and mutation to converge to the optimal solution.
\subsection{ Data preparation and population generation}
A genome is an expansion of DNA sequence which is a set of nucleotides that encodes a protein. These nucleotides are bounded together and each order define a specific protein. Three nucleotides form a triplet and interpreted as a specific amino acid ~\cite{Harvey00}. For this reason, the whole sequence is divided into a set of triplets.  As each DNA sequence contains a combination of four nucleobases\textemdash guanine (G), adenine (A), thymine (T), and cytosine (C), therefore we will have $4^3=64$ possible triplets. As mentioned earlier that the classical Huffman code requires reading the whole sequence prior to encoding. So, in this case, the whole DNA sequence is read considering different triplets as distinct symbols. Frequencies for each of the triplets are calculated and  the adjacency matrix (a symmetric 2D matrix) is also created which represents how many times a triplet occurs after each of the other triplets.  After that, a random set of Huffman trees are generated for the input symbols (triplets) which are considered as the initial population for the GA.
 \subsection{Genetic operations}
\subsubsection{Selection}

@@

\subsubsection{Crossover}

@@

\subsubsection{Mutation}
